WHOOMP! (There It Is) \\ Rapid Bayesian position reconstruction for gravitational\nobreakdashes-wave transients

### 2015-07-08 ms.tex
Add conclusion

With just one thread, the \ac{BAYESTAR} analysis takes 87\nobreakdashes--384~s, already orders of magnitude faster than the full parameter estimation. With 32 threads, \ac{BAYESTAR} takes just 4\nobreakdashes--15~s. This latency is comparable to the other stages (data aggregation, trigger generation, alert distribution) in the real\nobreakdashes-time \ac{BNS} analysis; therefore any significant further speedup would require significant changes to all of the analyses preceeding \ac{BAYESTAR}. The 32\nobreakdashes-thread configuration is representative of how \ac{BAYESTAR} might be deployed in early Advanced \ac{LIGO}.\footnote{\ac{BAYESTAR} has been successfully ported to the Intel's Many Integrated Core~(MIC) architecture and has been tested in a 500 thread configuration on a system with dual Intel Xeon Phi 5110P coprocessors.} For comparison, sky localization with LALINFERENCE takes about 100 hours \cite{BerryLocalization}.

@@

>>>>

With just one thread, the \ac{BAYESTAR} analysis takes 87\nobreakdashes--384~s, already orders of magnitude faster than the full parameter estimation. With 32 threads, \ac{BAYESTAR} takes just 4\nobreakdashes--15~s. In practice, \ac{BAYESTAR}'s data handling (reading the detectors' \acp{PSD}, communicating with the \ac{GW} candidate database, writing FITS files) take an additional 15~s or so, though this overhead could be reduced by parallelizing many of these steps. The overall latency is comparable to the other stages (data aggregation, trigger generation, alert distribution) in the real\nobreakdashes-time \ac{BNS} analysis; therefore any significant further speedup would require significant changes through Advanced \acs{LIGO} computing and infrastructure. The 32\nobreakdashes-thread configuration is representative of how \ac{BAYESTAR} might be deployed in early Advanced \ac{LIGO}.\footnote{\ac{BAYESTAR} has been successfully ported to the Intel's Many Integrated Core~(MIC) architecture and has been tested in a 500 thread configuration on a system with dual Intel Xeon Phi 5110P coprocessors.} For comparison, sky localization with LALINFERENCE takes about 100 hours \cite{BerryLocalization}.

@@

\section{Conclusion}

We have presented a novel, fast, accurate, Bayesian algorithm for inferring the sky locations of compact binary merger sources that may soon be detected by advanced ground\nobreakdashes-based \ac{GW} detectors. For \ac{BNS} systems with small spins, we have shown that \ac{BAYESTAR} produces sky maps that are as accurate as the full \ac{MCMC} parameter estimation code, but can do so within $\sim$10 seconds after a detection. Still faster response times should be possible in the future (if warranted) by deploying \ac{BAYESTAR} on machines with more cores, or by distributing \ac{BAYESTAR} across multiple computers.

Following a \ac{BNS} merger, the signal will be detected by the matched\nobreakdashes-filter pipeline within tens of seconds; an alert containing the time and estimated significance of the event can be distributed almost immediately (although a human validation stage that may be present at the beginning of the first observing run may introduce some additional latency). The localization from \ac{BAYESTAR} will be available tens of seconds to a minute later. Finally, the refined localiation and the detailed estimates of masses and spins from LALINFERENCE will be distributed hours to days later.

Relevant timescales for possible \ac{EM} counterparts to \ac{GW} signals include seconds (the prompt \ac{GRB} signature), hundreds of seconds (extended emission and X\nobreakdashes-ray plateaus that are observed for some short \acp{GRB}), minutes to hours (X\nobreakdashes-ray and optical afterglow), hours to days (the kilonova or the blue flashes associated with unbound ejecta or disk winds), and days to years (the radio afterglow). For the first time, we are able to provide accurate localizations before the peak of any of these \ac{EM} signatures (except for the short \ac{GRB} or any pre\nobreakdashes-merger signal). Even for components like the kilonova that should peak within hours to days, the availability of the localizations within seconds might provide a window of several hours to obtain tiled images of the area before the \ac{EM} emission begins. These could be used as reference images, crucial at optical wavelengths for establishing the rapid rise and quickly distinguishing from slower background transients.
